Symbolic Weighted Language Models and\\ Quantitative Parsing over Infinite Alphabets

### 2021-10-11 algo-best-alt.tex
parentheses in exemples VPA (linearization trees)

\subsection*{(Alternative Algorithm for best-search)}
We associate to the \SWVPA $A$ a $\Semiring$-labeled graph
$\G_A = \< V_A, E_A, \eta_A >$
with set of vertices $V_A = (Q \times Q) \cup (Q \times P \times Q)$,
set of edges $E_A = V_A \times V_A$
(an edge $\< v_1, v_2>$, with $v_1, v_2 \in V_A$ is denoted by $v_1 \to v_2$),
and with an edge labelling  
$\eta_A : E_A \to \Semiring$ defined as follows,
for $q_0, q_1, q_2, q_3 \in Q$, $p, p' \in P$:
%
\[
\begin{array}{rclcl}
\< q_0, q_1> & \to & \< q_0, q_2> & : &
\bigoplus_{\Deltai} \weiei(q_1, q_2)\\
%
\< q_1, p, q_2> & \to & \<q_0, q_3> & : &
\bigoplus_{\Deltac} \bigl( \weiec(q_0, p, q_1) \otimes \bigoplus^2_{\Deltar} \weir(q_2, p, q_3)\bigr)\\ 
%
\< q_0, q_2> & \to & \< q_0, q_3> & : & 
\bigoplus_{\Deltar} \weier(q_2, q_3)\\
% 
\< q_0, p, q_1> & \to & \< q_0, p, q_2> & : &
\bigoplus_{\Deltai} \weii(q_1, p, q_2)\\[1mm]
%
\< q_1, p', q_2> & \to & \< q_0, p, q_3> & : &
\bigoplus^2_{\Deltac} \bigl[ \weic(q_0, p, p', q_1) \otimes_2 {\bigoplus_{\Deltar}} \weir(q_2, p', q_3) \bigr]\\
\end{array}
\]
Removing from $\G_A$ every edge $e \in E_A$ such that $\eta(e) = \zero$ 
is preserving the results below.
%
A \emph{path} of $\G_A$ is a sequence $v_0,\ldots, v_n \in V^*_A$,
%for all $0 \leq i < n$. %the second component of $e_i$ is the first component of $e_{i+1}$.
%We write $\src(\pi) = e_1$ and $\trg(\pi) = e_n$.
it is assigned a weight value $\weight(\pi) = \bigotimes_{i=1}^{n} \eta(v_{i-1} \to v_i)$.

@@

We show now by induction on $n$ that we can construct a path $\pi$ of weight $b_\bot(q_1, q_2)$
from $\tau_0,\ldots, \tau_n$ and $w_0,\ldots, w_n$. ...\qed
\end{proof}

>>>>

\subsection*{(Alternative Algorithm for best-search)}
We shall compute the values of the functions $b_\bot$ and $b_\top$
by searching shortests paths in a $\Semiring$-labeled graph $\G_A$
associated to the \SWVPA $A$. 
This graph is defined by $\G_A = \< V_A, E_A, \eta_A >$, 
with set of vertices $V_A = (Q \times Q) \cup (Q \times P \times Q)$,
set of edges $E_A = V_A \times V_A$
(an edge $\< v_1, v_2>$, with $v_1, v_2 \in V_A$ is denoted by $v_1 \to v_2$),
and with an edge labelling  
$\eta_A : E_A \to \Semiring$ defined as follows,
for $q_0, q_1, q_2, q_3 \in Q$, $p, p' \in P$:
%
\[
\begin{array}{rclcl}
\< q_0, q_1> & \to & \< q_0, q_2> & : &
\bigoplus_{\Deltai} \weiei(q_1, q_2)\\
%
\< q_1, p, q_2> & \to & \<q_0, q_3> & : &
\bigoplus_{\Deltac} \bigl( \weiec(q_0, p, q_1) \otimes \bigoplus^2_{\Deltar} \weir(q_2, p, q_3)\bigr)\\ 
%
\< q_0, q_2> & \to & \< q_0, q_3> & : & 
\bigoplus_{\Deltar} \weier(q_2, q_3)\\
% 
\< q_0, p, q_1> & \to & \< q_0, p, q_2> & : &
\bigoplus_{\Deltai} \weii(q_1, p, q_2)\\[1mm]
%
\< q_1, p', q_2> & \to & \< q_0, p, q_3> & : &
\bigoplus^2_{\Deltac} \bigl[ \weic(q_0, p, p', q_1) \otimes_2 {\bigoplus_{\Deltar}} \weir(q_2, p', q_3) \bigr]\\
\end{array}
\]
Removing from $\G_A$ every edge $e \in E_A$ such that $\eta(e) = \zero$ 
is preserving the results below.
%
A \emph{path} of $\G_A$ is a sequence $v_0,\ldots, v_n \in V^*_A$,
%for all $0 \leq i < n$. %the second component of $e_i$ is the first component of $e_{i+1}$.
%We write $\src(\pi) = e_1$ and $\trg(\pi) = e_n$.
it is assigned a weight value $\weight(\pi) = \bigotimes_{i=1}^{n} \eta(v_{i-1} \to v_i)$.

@@

We show now the existence of a path $\pi$ as expected, by induction on $n$.
%that we can construct a path $\pi$ of weight $b_\bot(q_1, q_2)$
%from $\tau_0,\ldots, \tau_n$ and $w_0,\ldots, w_n$. ...\qed
\end{proof}

### 2021-10-11 main.tex
parentheses in exemples VPA (linearization trees)

%\subsection{Properties}
\medskip\noindent
Similarly to \VPA~\cite{AlurMadhusudan09nested}
and \SVPA~\cite{dAntonyAlur14SVPDA},
the class of \SWVPA is closed under the binary operators of the underlying semiring.
%
\begin{proposition}\label{prop:SWVPA-product}
Let $A_1$ and $A_2$ be two \SWVPA
over the same $\Delta$, commutative $\Semiring$ and effective $\bar\Phi$.
There exist two effectively constructible $\SWVPA$
$A_1 \oplus A_2$ and $A_1 \otimes A_2$,
such that for every $s \in \Delta^*$,
$(A_1 \oplus A_2)(s) = A_1(s) \oplus A_2(s)$ and
$(A_1 \otimes A_2)(s) = A_1(s) \otimes A_2(s)$.
\end{proposition}
%
\begin{proof}
We do a classical product construction. % see Appendix~\ref{sec:product} for details.
%similar to the case of the Boolean semiring~\cite{dAntonyAlur14SVPDA}.
\input{app-product}
\end{proof}
\noindent
%We present now a procedure for searching, for a~$\SWVPA$ $A$,
%a word %$w \in \Delta^*$
%of minimal weight for~$A$.
We present now a procedure for searching a word of minimal weight for a~$\SWVPA$~$A$.
%as stated in the following proposition. %\wrt~$\leq_\oplus$,
%It requires the following property of label theories.
%
\begin{proposition}\label{th:best-search}
For a \SWVPA $A$
over $\Delta$,
$\Semiring$ commutative, bounded, total and complete, %semiring
and $\bar\Phi$ effective, %and $k$-convex, % label theory,
one can construct in PTIME a word $t \in \Delta^*$
such that $A(t)$ is minimal \wrt the natural ordering $\leq_\oplus$ for $\Semiring$.
\end{proposition}
%\florent{total?}
%
Let $A = \< Q, P, \init, \bar\wei, \final >$.
%
We propose a method for %Dijkstra-like algorithm 
computing,
for every $q, q' \in Q$,
the minimum, \wrt $\leq_\oplus$, of the function
$\beta_{q, q'} : t \mapsto \weight_A(\config{q}{\bot}, t, \config{q'}{\bot})$.
Let us denote by $b_\bot(q, q')$ this minimum.
\florent{it exists by effectiveness of label th. ?}
%
By definition of $\leq_\oplus$, and since $\Semiring$ is total,
\reviews{1) $\Semiring$ total $\to$ $\Semiring$ extremal}
it holds that:
%
\begin{equation}\label{eq:bbot}
  b_\bot(q, q') = \bigoplus_{t\in \Delta^*}
  \textstyle
  \weight_A\bigl(\config{q}{\bot}, t, \config{q'}{\bot}\bigr).
\end{equation}
The infinite sum in~\eqref{eq:bbot} is well defined since $\Semiring$ is complete.
%
Following~\eqref{eq:SWVPA-value}, and the associativity, commutativity
and distributivity for~$\otimes$ and~$\oplus$, the minimum of~$A(t)$ is:
\begin{equation}\label{eq:min}
{\displaystyle\bigoplus_{t\in \Delta^*}} A(t)
=
{\displaystyle\bigoplus_{t\in \Delta^*}}
{\displaystyle\bigoplus_{q, q' \in Q}} \textstyle
\mathsf{in}(q) \mathop{\otimes}
\beta_{q, q'}(t)
\mathop{\otimes} \mathsf{out}(q')
=
{\displaystyle\bigoplus_{q, q' \in Q}} \textstyle
\mathsf{in}(q) \mathop{\otimes}
b_\bot(q, q')
\mathop{\otimes} \mathsf{out}(q')
\end{equation}

>>>>

%\subsection{Properties}
\medskip\noindent
Similarly to \VPA~\cite{AlurMadhusudan09nested}
and \SVPA~\cite{dAntonyAlur14SVPDA},
the class of \SWVPA is closed under the binary operators of the underlying semiring.
%
\begin{proposition}\label{prop:SWVPA-product}
Let $A_1$ and $A_2$ be two \SWVPA
over the same $\Delta$, commutative $\Semiring$ and effective $\bar\Phi$.
There exist two effectively constructible $\SWVPA$
$A_1 \oplus A_2$ and $A_1 \otimes A_2$,
such that for every $s \in \Delta^*$,
$(A_1 \oplus A_2)(s) = A_1(s) \oplus A_2(s)$ and
$(A_1 \otimes A_2)(s) = A_1(s) \otimes A_2(s)$.
\end{proposition}
%
\begin{proof}
%We do a classical product construction, see Appendix~\ref{sec:product} for details.
%similar to the case of the Boolean semiring~\cite{dAntonyAlur14SVPDA}.
\input{app-product}
\end{proof}
\noindent
%We present now a procedure for searching, for a~$\SWVPA$ $A$,
%a word %$w \in \Delta^*$
%of minimal weight for~$A$.
We present now a procedure for searching a word of minimal weight for a~$\SWVPA$~$A$.
%as stated in the following proposition. %\wrt~$\leq_\oplus$,
%It requires the following property of label theories.
%
\begin{proposition}\label{th:best-search}
For a \SWVPA $A$
over $\Delta$,
$\Semiring$ commutative, bounded, total and complete, %semiring
and $\bar\Phi$ effective, %and $k$-convex, % label theory,
one can construct in PTIME a word $t \in \Delta^*$
such that $A(t)$ is minimal \wrt the natural ordering $\leq_\oplus$ for $\Semiring$.
\end{proposition}
%\florent{total?}
%
Let $A = \< Q, P, \init, \bar\wei, \final >$.
%
We propose a method for %Dijkstra-like algorithm 
computing,
for every $q, q' \in Q$,
the minimum, \wrt $\leq_\oplus$, of the function
$\beta_{q, q'} : t \mapsto \weight_A(\config{q}{\bot}, t, \config{q'}{\bot})$.
Let us denote by $b_\bot(q, q')$ this minimum.
\florent{it exists by effectiveness of label th. ?}
%
By definition of $\leq_\oplus$, and since $\Semiring$ is total,
\reviews{1) $\Semiring$ total $\to$ $\Semiring$ extremal}
it holds that
(the infinite sum in~\eqref{eq:bbot} is well defined since $\Semiring$ is complete):
%
\begin{equation}\label{eq:bbot}
  b_\bot(q, q') = \bigoplus_{t\in \Delta^*}
  \textstyle
  \weight_A\bigl(\config{q}{\bot}, t, \config{q'}{\bot}\bigr).
\end{equation}
%
Following~\eqref{eq:SWVPA-value}, and the associativity, commutativity
and distributivity for~$\otimes$ and~$\oplus$, the minimum of~$A(t)$ is:
\begin{equation}\label{eq:min}
{\displaystyle\bigoplus_{t\in \Delta^*}} A(t)
=
{\displaystyle\bigoplus_{t\in \Delta^*}}
{\displaystyle\bigoplus_{q, q' \in Q}} \textstyle
\mathsf{in}(q) \mathop{\otimes}
\beta_{q, q'}(t)
\mathop{\otimes} \mathsf{out}(q')
=
{\displaystyle\bigoplus_{q, q' \in Q}} \textstyle
\mathsf{in}(q) \mathop{\otimes}
b_\bot(q, q')
\mathop{\otimes} \mathsf{out}(q')
\end{equation}
